---
title: 'Lab02: Polynomial Regression'
author: "Mehmet Gönen"
date: "October 1, 2018"
output: html_document
runtime: shiny
---

## Importing Data
```{r}
# read data into memory
data_set <- read.csv("lab02_data_set.csv")

# get x and y values
x <- data_set$x
y <- data_set$y

# get number of samples
N <- length(y)

x_test <- seq(from = 0, to = 60, by = 0.1)
```

## Plotting Data
```{r}
# plot the training data to see the trend
plot(x, y, las = 1)
```

## Linear Regression
\begin{align*}
\underbrace{\begin{bmatrix} N & \sum\limits_{i = 1}^{N} x_{i}\\ \sum\limits_{i = 1}^{N} x_{i} & \sum\limits_{i = 1}^{N} x_{i}^{2} \end{bmatrix}}_{\mathbf{A}} \underbrace{\begin{bmatrix}w_{0} \\ w_{1} \end{bmatrix}}_{\boldsymbol{w}} &= \underbrace{\begin{bmatrix}\sum\limits_{i = 1}^{N} y_{i} \\ \sum\limits_{i = 1}^{N} y_{i} x_{i} \end{bmatrix}}_{\boldsymbol{b}} \\
\boldsymbol{w} &= \mathbf{A}^{-1} \boldsymbol{b}
\end{align*}

```{r}
# calculate A
A <- matrix(c(N, sum(x), sum(x), sum(x^2)), 2, 2)
print(A)

# calculate b
b <- c(sum(y), sum(y * x))
print(b)

# calculate w
w <- chol2inv(chol(A)) %*% b
print(w)
```

```{r}
plot(x, y, las = 1)
points(x_test, cbind(1, x_test) %*% w, lwd = 2, col = "blue", type = "l")
```

## Polynomial Regression
\begin{align*}
\underbrace{\begin{bmatrix} N & \sum\limits_{i = 1}^{N} x_{i} & \dots &  \sum\limits_{i = 1}^{N} x_{i}^{K}\\ \sum\limits_{i = 1}^{N} x_{i} & \sum\limits_{i = 1}^{N} x_{i}^{2} & \dots & \sum\limits_{i = 1}^{N} x_{i}^{K + 1}\\ \vdots & \vdots & \ddots & \vdots \\
\sum\limits_{i = 1}^{N} x_{i}^{K} & \sum\limits_{i = 1}^{N} x_{i}^{K + 1} & \dots & \sum\limits_{i = 1}^{N} x_{i}^{2K}\end{bmatrix}}_{\mathbf{A}} \underbrace{\begin{bmatrix}w_{0} \\ w_{1} \\ \vdots \\ w_{K} \end{bmatrix}}_{\boldsymbol{w}} &= \underbrace{\begin{bmatrix}\sum\limits_{i = 1}^{N} y_{i} \\ \sum\limits_{i = 1}^{N} y_{i} x_{i} \\ \vdots \\ \sum\limits_{i = 1}^{N} y_{i} x_{i}^{K} \end{bmatrix}}_{\boldsymbol{b}} \\
\boldsymbol{w} &= \mathbf{A}^{-1} \boldsymbol{b}
\end{align*}

```{r}
polynomial_regression <- function(x, y, K) {
  # calculate A and b
  A <- matrix(0, K + 1, K + 1)
  for (i in 0:K) {
    for (j in 0:K) {
      A[i + 1, j + 1] <- sum(x^i * x^j)
    }
  }
  
  b <- rep(0, K + 1)
  for (i in 0:K) {
    b[i + 1] <- sum(y * x^i)
  }
  
  # calculate w
  w <- chol2inv(chol(A)) %*% b
  
  return(w)
}
```

```{r}
inputPanel(
  sliderInput("polynomial_degree1", label = "Polynomial degree:", min = 0, max = 10, step = 1, value = 1)
)

renderPlot({
  K <- as.numeric(input$polynomial_degree1)
  w <- polynomial_regression(x, y, K)
  plot(x, y, las = 1)
  points(x_test, sapply(0:K, function(k) x_test^k) %*% w, lwd = 2, col = "blue", type = "l")
}, width = 800, height = 600)
```

\begin{align*}
\begin{bmatrix} N & \sum\limits_{i = 1}^{N} x_{i} & \dots &  \sum\limits_{i = 1}^{N} x_{i}^{K}\\ \sum\limits_{i = 1}^{N} x_{i} & \sum\limits_{i = 1}^{N} x_{i}^{2} & \dots & \sum\limits_{i = 1}^{N} x_{i}^{K + 1}\\ \vdots & \vdots & \ddots & \vdots \\
\sum\limits_{i = 1}^{N} x_{i}^{K} & \sum\limits_{i = 1}^{N} x_{i}^{K + 1} & \dots & \sum\limits_{i = 1}^{N} x_{i}^{2K}\end{bmatrix} \begin{bmatrix}w_{0} \\ w_{1} \\ \vdots \\ w_{K} \end{bmatrix} &= \begin{bmatrix}\sum\limits_{i = 1}^{N} y_{i} \\ \sum\limits_{i = 1}^{N} y_{i} x_{i} \\ \vdots \\ \sum\limits_{i = 1}^{N} y_{i} x_{i}^{K} \end{bmatrix} \\
\\
\underbrace{\begin{bmatrix}1 & 1 & \dots & 1\\ x_{1} & x_{2} & \dots & x_{N} \\ \vdots & \vdots & \ddots & \vdots \\ x_{1}^{K} & x_{2}^{K} & \dots & x_{N}^{K} \end{bmatrix}}_{\mathbf{D}^{\top}} \underbrace{\begin{bmatrix}1 & x_{1} & \dots & x_{1}^{K}\\ 1 & x_{2} & \dots & x_{2}^{K} \\ \vdots & \vdots & \ddots & \vdots \\ 1 & x_{N} & \dots & x_{N}^{K} \end{bmatrix}}_{\mathbf{D}} \underbrace{\begin{bmatrix}w_{0} \\ w_{1} \\ \vdots \\ w_{K} \end{bmatrix}}_{\boldsymbol{w}} &= \underbrace{\begin{bmatrix}1 & 1 & \dots & 1\\ x_{1} & x_{2} & \dots & x_{N} \\ \vdots & \vdots & \ddots & \vdots \\ x_{1}^{K} & x_{2}^{K} & \dots & x_{N}^{K} \end{bmatrix}}_{\mathbf{D}^{\top}} \underbrace{\begin{bmatrix}y_{1} \\ y_{2} \\ \vdots \\ y_{N} \end{bmatrix}}_{\boldsymbol{y}} \\
\boldsymbol{w} &= \left(\mathbf{D}^{\top}\mathbf{D}\right)^{-1} \mathbf{D}^{\top} \boldsymbol{y}
\end{align*}

```{r}
polynomial_regression <- function(x, y, K) {
  # calculate D
  D <- sapply(0:K, function(k) x^k)
  
  # calculate w
  w <- chol2inv(chol(t(D) %*% D)) %*% (t(D) %*% y)
  
  return(w)
}
```

```{r}
inputPanel(
  sliderInput("polynomial_degree2", label = "Polynomial degree:", min = 0, max = 10, step = 1, value = 1)
)

renderPlot({
  K <- as.numeric(input$polynomial_degree2)
  w <- polynomial_regression(x, y, K)
  plot(x, y, las = 1)
  points(x_test, sapply(0:K, function(k) x_test^k) %*% w, lwd = 2, col = "blue", type = "l")
}, width = 800, height = 600)
```

## Nonlinear Regression
```{r}
nonlinear_regression <- function(x, y, centers, sigma) {
  # calculate D
  D <- cbind(1, sapply(centers, function(center) exp(-(x - center)^2 / (2 * sigma^2))))
  
  # calculate w
  w <- chol2inv(chol(t(D) %*% D)) %*% (t(D) %*% y)
  
  return(w)
}
```

```{r}
inputPanel(
  sliderInput("density_sigma", label = "Sigma:", min = 1, max = 10, step = 1, value = 5)
)

renderPlot({
  centers <- seq(from = 5, to = 55, by = 5)
  sigma <- as.numeric(input$density_sigma)
  w <- nonlinear_regression(x, y, centers, sigma)
  D_test <- cbind(1, sapply(centers, function(center) exp(-(x_test - center)^2 / (2 * sigma^2))))
  
  plot(x, y, las = 1)
  points(centers, rep(0, length(centers)), pch = 19, cex = 1.5, col = "red")
  points(x_test, D_test %*% w, lwd = 2, col = "blue", type = "l")
}, width = 800, height = 600)
```