---
title: 'Lab04: Linear Discrimination'
author: "Mehmet Gönen"
date: "October 14, 2019"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(MASS)
safelog <- function(x) {
  return (log(x + 1e-100))
}
```

## Parameters
```{r}
set.seed(421)
# mean parameters
class_means <- matrix(c(+0.0, +1.5, -2.5, -3.0, +2.5, -3.0), 2, 3)
# covariance parameters
class_covariances <- array(c(+1.0, +0.2, +0.2, +3.2,
                             +1.6, -0.8, -0.8, +1.0,
                             +1.6, +0.8, +0.8, +1.0), c(2, 2, 3))
# sample sizes
class_sizes <- c(100, 100, 100)
```

## Data Generation
```{r}
# generate random samples
points1 <- mvrnorm(n = class_sizes[1], mu = class_means[,1], Sigma = class_covariances[,,1])
points2 <- mvrnorm(n = class_sizes[2], mu = class_means[,2], Sigma = class_covariances[,,2])
points3 <- mvrnorm(n = class_sizes[3], mu = class_means[,3], Sigma = class_covariances[,,3])
X <- rbind(points1, points2, points3)
colnames(X) <- c("x1", "x2")

# generate corresponding labels
y <- c(rep(1, class_sizes[1]), rep(2, class_sizes[2]), rep(3, class_sizes[3]))
```

## Exporting Data
```{r}
# write data to a file
write.csv(x = cbind(X, y), file = "lab04_data_set.csv", row.names = FALSE)
```

## Plotting Data
```{r, fig.height = 8, fig.width = 8, fig.keep = "all"}
# plot data points generated
plot(points1[,1], points1[,2], type = "p", pch = 19, col = "red", las = 1,
     xlim = c(-6, 6), ylim = c(-6, 6),
     xlab = "x1", ylab = "x2")
points(points2[,1], points2[,2], type = "p", pch = 19, col = "green")
points(points3[,1], points3[,2], type = "p", pch = 19, col = "blue")
```

## Importing Data
```{r}
# read data into memory
data_set <- read.csv("lab04_data_set.csv")

# get X and y values
X <- cbind(data_set$x1, data_set$x2)
y_truth <- data_set$y

# get number of classes and number of samples
K <- max(y_truth)
N <- length(y_truth)

# one-of-K-encoding
Y_truth <- matrix(0, N, K)
Y_truth[cbind(1:N, y_truth)] <- 1
```

## Softmax Function
$\textrm{softmax}_{c}(\boldsymbol{w}_{c}^{\top} \boldsymbol{x} + w_{c0}) = \dfrac{\exp\left[(\boldsymbol{w}_{c}^{\top} \boldsymbol{x} + w_{c0})\right]}{\sum\limits_{k=1}^{K}\exp\left[(\boldsymbol{w}_{k}^{\top} \boldsymbol{x} + w_{k0})\right]}$

```{r}
# define the softmax function
softmax <- function(X, W, w0) {
  scores <- cbind(X, 1) %*% rbind(W, w0)
  scores <- exp(scores - matrix(apply(scores, MARGIN = 1, FUN = max), nrow = nrow(scores), ncol = ncol(scores), byrow = FALSE))
  scores <- scores / matrix(rowSums(scores), nrow(scores), ncol(scores), byrow = FALSE)
  return (scores)
}
```

## Gradient Functions
\begin{align*}
\dfrac{\partial \textrm{Error}}{\partial \boldsymbol{w}_{c}} &= -\sum\limits_{i = 1}^{N} (y_{ic} - \widehat{y}_{ic})\boldsymbol{x}_{i} \\
\dfrac{\partial \textrm{Error}}{\partial w_{c0}} &= -\sum\limits_{i = 1}^{N} (y_{ic} - \widehat{y}_{ic}) 
\end{align*}

```{r}
# define the gradient functions
gradient_W <- function(X, Y_truth, Y_predicted) {
  return (-sapply(X = 1:ncol(Y_truth), function(c) colSums(matrix(Y_truth[,c] - Y_predicted[,c], nrow = nrow(X), ncol = ncol(X), byrow = FALSE) * X)))
}

gradient_w0 <- function(Y_truth, Y_predicted) {
  return (-colSums(Y_truth - Y_predicted))
}
```

## Algorithm Parameters
```{r}
# set learning parameters
eta <- 0.01
epsilon <- 1e-3
```

## Parameter Initialization
```{r}
# randomly initalize w and w0
set.seed(421)
W <- matrix(runif(ncol(X) * K, min = -0.01, max = 0.01), ncol(X), K)
w0 <- runif(K, min = -0.01, max = 0.01)
```

## Iterative Algorithm 
$\textrm{Error} = -\sum\limits_{i = 1}^{N} \sum\limits_{c = 1}^{K} \left[ y_{ic}\log(\widehat{y}_{ic}) \right]$
```{r}
# learn W and w0 using gradient descent
iteration <- 1
objective_values <- c()
while (1) {
  Y_predicted <- softmax(X, W, w0)
  
  objective_values <- c(objective_values, -sum(Y_truth * safelog(Y_predicted)))
  
  W_old <- W
  w0_old <- w0
  
  W <- W - eta * gradient_W(X, Y_truth, Y_predicted)
  w0 <- w0 - eta * gradient_w0(Y_truth, Y_predicted)
  
  if (sqrt(sum((w0 - w0_old)^2) + sum((W - W_old)^2)) < epsilon) {
    break
  }
  
  iteration <- iteration + 1
}
print(W)
print(w0)
```

## Convergence
```{r, fig.height = 6, fig.width = 8}
# plot objective function during iterations
plot(1:iteration, objective_values,
     type = "l", lwd = 2, las = 1,
     xlab = "Iteration", ylab = "Error")
```

## Training Performance
```{r}
# calculate confusion matrix
y_predicted <- apply(Y_predicted, MARGIN = 1, FUN = which.max)
confusion_matrix <- table(y_predicted, y_truth)
print(confusion_matrix)
```

## Visualization
```{r, fig.height = 8, fig.width = 8, fig.keep = "all"}
# evaluate discriminat function on a grid
x1_interval <- seq(from = -6, to = +6, by = 0.06)
x2_interval <- seq(from = -6, to = +6, by = 0.06)
x1_grid <- matrix(x1_interval, nrow = length(x1_interval), ncol = length(x1_interval), byrow = FALSE)
x2_grid <- matrix(x2_interval, nrow = length(x2_interval), ncol = length(x2_interval), byrow = TRUE)

discriminant_values <- array(0, c(length(x1_interval), length(x2_interval), K))
for (c in 1:K) {
  f <- function(x1, x2) {W[1, c] * x1 + W[2, c] * x2 + w0[c]}
  discriminant_values[,,c] <- exp(matrix(mapply(f, x1_grid, x2_grid), nrow(x2_grid), ncol(x2_grid)))
}

plot(X[y_truth == 1, 1], X[y_truth == 1, 2], type = "p", pch = 19, col = "red",
     xlim = c(-6, +6),
     ylim = c(-6, +6),
     xlab = "x1", ylab = "x2", las = 1)
points(X[y_truth == 2, 1], X[y_truth == 2, 2], type = "p", pch = 19, col = "green")
points(X[y_truth == 3, 1], X[y_truth == 3, 2], type = "p", pch = 19, col = "blue")
points(x1_grid[apply(discriminant_values, c(1, 2), which.max) == 1], x2_grid[apply(discriminant_values, c(1, 2), which.max) == 1], col = rgb(red = 1, green = 0, blue = 0, alpha = 0.05), pch = 16)
points(x1_grid[apply(discriminant_values, c(1, 2), which.max) == 2], x2_grid[apply(discriminant_values, c(1, 2), which.max) == 2], col = rgb(red = 0, green = 1, blue = 0, alpha = 0.05), pch = 16)
points(x1_grid[apply(discriminant_values, c(1, 2), which.max) == 3], x2_grid[apply(discriminant_values, c(1, 2), which.max) == 3], col = rgb(red = 0, green = 0, blue = 1, alpha = 0.05), pch = 16)
A <- discriminant_values[,,1]
B <- discriminant_values[,,2]
C <- discriminant_values[,,3]
A[A < B & A < C] <- NA
B[B < A & B < C] <- NA
C[C < A & C < B] <- NA
discriminant_values[,,1] <- A
discriminant_values[,,2] <- B
discriminant_values[,,3] <- C
contour(x1_interval, x1_interval, discriminant_values[,,1] - discriminant_values[,,2], levels = c(0), add = TRUE, lwd = 2, drawlabels = FALSE)
contour(x1_interval, x1_interval, discriminant_values[,,1] - discriminant_values[,,3], levels = c(0), add = TRUE, lwd = 2, drawlabels = FALSE)
contour(x1_interval, x1_interval, discriminant_values[,,2] - discriminant_values[,,3], levels = c(0), add = TRUE, lwd = 2, drawlabels = FALSE)
points(X[y_predicted != y_truth, 1], X[y_predicted != y_truth, 2], cex = 1.5, lwd = 2)
```