---
title: 'Lab06: Nonparametric Methods'
author: "Mehmet Gönen"
date: "November 5, 2018"
output: html_document
runtime: shiny
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Parameters
```{r}
set.seed(421)
# mean parameters
class_means <- c(-2, +1, 3)
# standard deviation parameters
class_deviations <- c(2.0, 1.2, 1.6)
# sample sizes
class_sizes <- c(20, 15, 25)
```

## Data Generation
```{r}
# generate random samples
points1 <- rnorm(n = class_sizes[1], mean = class_means[1], sd = class_deviations[1])
points2 <- rnorm(n = class_sizes[2], mean = class_means[2], sd = class_deviations[2])
points3 <- rnorm(n = class_sizes[3], mean = class_means[3], sd = class_deviations[3])
x <- c(points1, points2, points3)

# generate corresponding labels
y <- c(rep(1, class_sizes[1]), rep(2, class_sizes[2]), rep(3, class_sizes[3]))
```

## Exporting Data
```{r}
# write data to a file
write.csv(x = cbind(x, y), file = "lab06_data_set.csv", row.names = FALSE)
```

## Plotting Data
```{r}
# plot densities used and data points generated together
renderPlot({
  data_interval <- seq(from = -8, to = +8, by = 0.01)
  density1 <- dnorm(data_interval, mean = class_means[1], sd = class_deviations[1])
  density2 <- dnorm(data_interval, mean = class_means[2], sd = class_deviations[2])
  density3 <- dnorm(data_interval, mean = class_means[3], sd = class_deviations[3])
  plot(data_interval, density1, type = "l", col = "red", lwd = 2, 
       ylim = c(-0.03, max(density1, density2, density3)),
       xlab = "x", ylab = "density", las = 1)
  points(data_interval, density2, type = "l", col = "green", lwd = 2)
  points(data_interval, density3, type = "l", col = "blue", lwd = 2)
  points(points1, rep(-0.01, class_sizes[1]), type = "p", pch = 19, col = "red")
  points(points2, rep(-0.02, class_sizes[2]), type = "p", pch = 19, col = "green")
  points(points3, rep(-0.03, class_sizes[3]), type = "p", pch = 19, col = "blue")
}, width = 800, height = 400)
```

## Importing Data
```{r}
# read data into memory
data_set <- read.csv("lab06_data_set.csv")

# get x and y values
x_train <- data_set$x
y_train <- data_set$y

# get number of classes and number of samples
K <- max(y_train)
N <- length(y_train)
```

## Drawing Parameters
```{r}
point_colors <- c("red", "green", "blue")
minimum_value <- -8
maximum_value <- +8
data_interval <- seq(from = minimum_value, to = maximum_value, by = 0.01)
```

## Histogram Estimator
$$\widehat{p}(x) = \dfrac{\#\{x_{i} \textrm{ is in the same bin as } x\}}{Nh}$$
```{r}
inputPanel(
  sliderInput("bin_width1", label = "Bin width:", min = 0.1, max = 2, step = 0.1, value = 0.5, animate = TRUE)
)

renderPlot({
  bin_width <- as.numeric(input$bin_width1)
  left_borders <- seq(from = minimum_value, to = maximum_value - bin_width, by = bin_width)
  right_borders <- seq(from = minimum_value + bin_width, to = maximum_value, by = bin_width)
  p_head <- sapply(1:length(left_borders), function(b) {sum(left_borders[b] < x_train & x_train <= right_borders[b])}) / (N * bin_width)

  plot(x_train, -0.01 * y_train, type = "p", pch = 19, col = point_colors[y_train],
       ylim = c(-0.03, max(p_head)), xlim = c(minimum_value, maximum_value),
       ylab = "density", xlab = "x", las = 1, main = sprintf("h = %g", bin_width))
  for (b in 1:length(left_borders)) {
    lines(c(left_borders[b], right_borders[b]), c(p_head[b], p_head[b]), lwd = 2, col = "black")
    if (b < length(left_borders)) {
      lines(c(right_borders[b], right_borders[b]), c(p_head[b], p_head[b + 1]), lwd = 2, col = "black") 
    }
  }
}, width = 800, height = 400)
```

## Naive Estimator
$$\widehat{p}(x) = \dfrac{\#\{x - h/2 \lt x_{i} \leq x + h/2\}}{Nh}$$
```{r}
inputPanel(
  sliderInput("bin_width2", label = "Bin width:", min = 0.1, max = 2, step = 0.1, value = 0.5, animate = TRUE)
)

renderPlot({
  bin_width <- as.numeric(input$bin_width2)
  p_head <- sapply(data_interval, function(x) {sum((x - 0.5 * bin_width) < x_train & x_train <= (x + 0.5 * bin_width))}) / (N * bin_width)

  plot(x_train, -0.01 * y_train, type = "p", pch = 19, col = point_colors[y_train],
       ylim = c(-0.03, max(p_head)), xlim = c(minimum_value, maximum_value),
       ylab = "density", xlab = "x", las = 1, main = sprintf("h = %g", bin_width))
  lines(data_interval, p_head, type = "l", lwd = 2, col = "black")
}, width = 800, height = 400)
```

## Kernel Estimator
\begin{align*}
\widehat{p}(x) &= \dfrac{1}{Nh} \sum \limits_{i = 1}^{N} K\left(\dfrac{x - x_{i}}{h}\right) \\
K(u) &= \dfrac{1}{\sqrt{2\pi}} \exp\left(-\dfrac{u^2}{2}\right)
\end{align*}
```{r}
inputPanel(
  sliderInput("bin_width3", label = "Bin width:", min = 0.1, max = 2, step = 0.1, value = 0.5, animate = TRUE)
)

renderPlot({
  bin_width <- as.numeric(input$bin_width3)
  p_head <- sapply(data_interval, function(x) {sum(1 / sqrt(2 * pi) * exp(-0.5 * (x - x_train)^2 / bin_width^2))}) / (N * bin_width)

  plot(x_train, -0.01 * y_train, type = "p", pch = 19, col = point_colors[y_train],
       ylim = c(-0.03, max(p_head)), xlim = c(minimum_value, maximum_value),
       ylab = "density", xlab = "x", las = 1, main = sprintf("h = %g", bin_width))
  lines(data_interval, p_head, type = "l", lwd = 2, col = "black")
}, width = 800, height = 400)
```

# k-Nearest Neighbor Estimator
\begin{align*}
\widehat{p}(x) &= \dfrac{k}{2Nd_{k}(x)} \\
d_{k}(x) &= \textrm{distance to k$^{th}$ closest data point to $x$} 
\end{align*}
```{r}
inputPanel(
  sliderInput("k1", label = "Neighborhood size (k):", min = 3, max = 19, step = 2, value = 11, animate = TRUE)
)

renderPlot({
  k <- as.numeric(input$k1)
  p_head <- sapply(data_interval, function(x) {k / (2 * N * sort(abs(x - x_train), decreasing = FALSE)[k])})

  plot(x_train, -0.01 * y_train, type = "p", pch = 19, col = point_colors[y_train],
       ylim = c(-0.03, max(p_head)), xlim = c(minimum_value, maximum_value),
       ylab = "density", xlab = "x", las = 1, main = sprintf("k = %g", k))
  lines(data_interval, p_head, type = "l", lwd = 2, col = "black")
}, width = 800, height = 400)
```

# Nonparametric Classification
$$g_{c}(x) = \dfrac{1}{Nh^{D}} \sum \limits_{i = 1}^{N} \left[K\left(\dfrac{x - x_{i}}{h}\right)y_{ic}\right]$$
```{r}
inputPanel(
  sliderInput("bin_width4", label = "Bin width:", min = 0.1, max = 2, step = 0.1, value = 0.5, animate = TRUE)
)

renderPlot({
  bin_width <- as.numeric(input$bin_width4)
  p_head <- matrix(0, length(data_interval), K)
  for (c in 1:K) {
    p_head[,c] <- sapply(data_interval, function(x) {sum(1 / sqrt(2 * pi) * exp(-0.5 * (x - x_train[y_train == c])^2 / bin_width^2))}) / (N * bin_width) 
  }

  plot(x_train, -0.01 * y_train, type = "p", pch = 19, col = point_colors[y_train],
     ylim = c(-0.03, max(p_head)), xlim = c(minimum_value, maximum_value),
     ylab = "density", xlab = "x", las = 1, main = sprintf("h = %g", bin_width))
  for (c in 1:K) {
    lines(data_interval, p_head[,c], type = "l", lwd = 2, col = point_colors[c]) 
  }
  points(data_interval, rep(max(p_head), length(data_interval)),
         col = point_colors[apply(p_head, 1, which.max)], pch = 19)
}, width = 800, height = 400)
```

# kNN Classification
$$g_{c}(x) = \dfrac{k_{c}}{k}$$
```{r}
inputPanel(
  sliderInput("k2", label = "Neighborhood size (k):", min = 3, max = 19, step = 2, value = 11, animate = TRUE)
)

renderPlot({
  k <- as.numeric(input$k2)
  p_head <- matrix(0, length(data_interval), K)
  for (c in 1:K) {
    p_head[,c] <- sapply(data_interval, function(x) {sum(y_train[order(abs(x - x_train), decreasing = FALSE)[1:k]] == c) / k}) 
  }

  plot(x_train, -0.01 * y_train, type = "p", pch = 19, col = point_colors[y_train],
       ylim = c(-0.03, max(p_head)), xlim = c(minimum_value, maximum_value),
       ylab = "density", xlab = "x", las = 1, main = sprintf("k = %g", k))
  for (c in 1:K) {
    lines(data_interval, p_head[,c], type = "l", lwd = 2, col = point_colors[c]) 
  }
  points(data_interval, rep(max(p_head), length(data_interval)),
         col = point_colors[apply(p_head, 1, which.max)], pch = 19)
}, width = 800, height = 400)
```